Canonical rings of rational divisors on projective space and Hirzebruch surfaces

### 2015-07-16 surfaces-write-up/surface-write-up.tex

\DeclareMathOperator{\supp}{Supp}
\DeclareMathOperator{\initial}{in_\prec}
\DeclareMathOperator{\gin}{gin}
\DeclareMathOperator{\Eff}{Eff}
\DeclareMathOperator{\sat}{sat}
\DeclareMathOperator{\newspan}{span}
\DeclareMathOperator{\proj}{Proj}
\DeclareMathOperator{\spec}{Spec}
\DeclareMathOperator{\Te}{T_=}
\DeclareMathOperator{\Tp}{T_+}
\DeclareMathOperator{\Tm}{T_-}
\DeclareMathOperator{\num}{p}
\DeclareMathOperator{\den}{q}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\Pic}{Pic}
%\captionsetup[table]{belowskip = 4pt}

@@

\ssec{Notation}
\begin{convention}
Notate a fractional (weil) divisor
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \di X \otimes \bq.
\end{align*}
where $\alpha_i \in \bq$ and $D_i \in \di X$ is an irreducible divisor.
\end{convention}

@@

\noindent
Suppose $e_1, \ldots, e_t \in \Sigma$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ are a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{t} \pdeg_i$
. Furthermore, every element $\sigma \in \Sigma$ can be written in
a canonical form $\lambda + \sum_{i=1}^{t} w_i e_i,$ with $\lambda = \sum_{i=1}^{r}s_i e_i$ with $0 \leq s_i < 1,$ the first
coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i,$ and $w_1,
\ldots, w_t \in \bz_{\geq 0}$.
\end{prop}

\begin{proof}
By assumption, $\sigma \in \Sigma$ can be written as $\sigma = \sum_{i=1}^{t} r_i e_i$ with $r_i \in \br$. Let $\{r\} := r - \lfloor r \rfloor$ denote the fractional part of $r$. Let $\lambda = \sum_{i=1}^{t}\{r_i\}$. Whence, we can write $\sigma = \lambda + \sum_{i=1}^{t}\lfloor r_i \rfloor e_i.$ Consequently, $\sigma$ lies in the $\bz_{\geq 0}$ span of $\lambda, e_1,\ldots, e_t$, which all have first coordinate less than $\sum_{i=1}^{t}r_i$. Ergo, $\Sigma$ is generated by elements whose first coordinate is less than $\sum_{i=1}^{t}r_i$.
\end{proof}

@@

Let $M$ be the set of monomials of $\beta_j$.  Then for $\mu\in M$, the corresponding lattice point is of the form
\[
	E_\mu + \sum_{k\in \Te} r_k e_k + \sum_{\substack{
	h \in \Tp \\ k \in \Tm}} r_{h,k} e_{h,k}
\]
where $E_\mu$ is a $\mathbb{R}_{\ge 0}$-combination of elements in
\begin{align*}
			& \{e_j : j\in \Te \cap \{1, 2, 3, 4\}\} \\
	\cup \; 	& \{e_{j, k}: j\in \Te \cap \{1, 2, 3, 4\}, k\in \Tm\}\\
	\cup \; 	& \{e_{k,j}: k\in \Tp, j\in \Tm \cap \{1, 2, 3, 4\}\}.
\end{align*}

Let $F\in R_D$ correspond to $E + \sum_{k\in \Te} (r_k) e_k + \sum_{
\substack{h \in \Tp \\ k \in \Tm}} (r_{h,k}) e_{h,k}$ where $(r)$
 signifies the fractional part of $r$.  Similarly, let $F_\mu \in
R_D$ correspond to $E_\mu + \sum_{\substack{h \in \Tp \\ k\in \Tm}}
(r_{h,k}) e_{h,k}$. Then $F - \sum_{\mu\in M} F_\mu = 0$ gives us a
relation in degree at most
\[
	\sigma + \deg(F)
\]

\noindent
which divides $(z_i - \beta_i) \prod_{i=1}^n y_i^{c_i}$.
We can bound $\deg(F)$ by
\[
	\deg(F) \le \left\{ \begin{split}
	\ell_i \gcd(\bida_i, \bidb_i)	\mbox{ if } i \in S_= \\
	\max_{j \in S_-}(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i))
	\mbox{ if } i \in S_+ \\
	\max_{j \in S_+}(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j))
	\mbox{ if } i \in S_- \end{split}  \right.
\]

\noindent
Therefore, $\ker(\psi)$ is generated in degree at most
\[
	\tau = \sigma
	+ \max \left(\max_{i\in S_=}(\ell_i \gcd(a_i, b_i)), \max_{\substack{
	i \in S_+ \\ j \in S_-}} (\ell_{i, j} (\bida_i \bidb_j - \bida_j
	\bidb_i)) \right).
\]
\end{proof}

@@

\noindent
where
\[
	\tau = \sigma
	+ \max( \max_{i\in S_=}(\ell_i \gcd(a_i, b_i)), \max_{\substack{
	i \in S_+ \\ j \in S_-}}(\ell_{i, j} (\bida_i \bidb_j - \bida_j
	\bidb_i))).
\]
\end{thm}

>>>>

\DeclareMathOperator{\fr}{frac}
\DeclareMathOperator{\supp}{Supp}
\DeclareMathOperator{\initial}{in_\prec}
\DeclareMathOperator{\gin}{gin}
\DeclareMathOperator{\Eff}{Eff}
\DeclareMathOperator{\sat}{sat}
\DeclareMathOperator{\newspan}{span}
\DeclareMathOperator{\proj}{Proj}
\DeclareMathOperator{\spec}{Spec}
\DeclareMathOperator{\Te}{T_=}
\DeclareMathOperator{\Tp}{T_+}
\DeclareMathOperator{\Tm}{T_-}
\DeclareMathOperator{\num}{p}
\DeclareMathOperator{\den}{q}
\DeclareMathOperator{\lcm}{lcm}
\DeclareMathOperator{\Pic}{Pic}
%\captionsetup[table]{belowskip = 4pt}

@@

\ssec{Notation}
\begin{convention}
Notate a fractional (Weil) divisor
\begin{align*}
	D = \sum_{i=1}^{n}\alpha_i D_i \in \di X \otimes \bq.
\end{align*}
where $\alpha_i \in \bq$ and $D_i \in \di X$ is an irreducible divisor.
\end{convention}

@@

\noindent
Suppose $e_1, \ldots, e_t \in \Sigma$ with $e_i = (\pdeg_i, c_1^i,
\ldots, c_n^i)$ are a set of {\bf extremal rays} of $\Sigma$,
meaning that the $\Sigma$ is contained in the $R_{\geq 0}$ span of
$e_1, \ldots, e_t$.  Then, as a semigroup, $\Sigma$ is generated by
elements whose first coordinate is less than $\sum_{i = 1}^{t}
\pdeg_i$. Furthermore, every element $\sigma \in \Sigma$ can be
written in a canonical form $\lambda + \sum_{i = 1}^{t} w_i e_i$ 
with $\lambda = \sum_{i = 1}^{r} s_i e_i$ with $0 \leq s_i < 1$, the
first coordinate of $\lambda$ less than $\sum_{i=1}^{t}\pdeg_i$,
and $w_1, \ldots, w_t \in \bz_{\geq 0}$.
\end{prop}

\begin{proof}
By assumption, $\sigma \in \Sigma$ can be written as $\sigma = \sum_
{i = 1}^{t} r_i e_i$ with $r_i \in \br$. Let $\fr(r) := r - \lfloor r
\rfloor$ denote the fractional part of $r$. Let $\lambda = \sum_{i = 1}
^{t} \fr(r_i)$. Whence, we can write $\sigma = \lambda + \sum_{i = 1}
^{t} \lfloor r_i \rfloor e_i.$ Consequently, $\sigma$ lies in the
$\bz_{\geq 0}$ span of $\lambda, e_1, \ldots, e_t$, which all have
first coordinate less than $\sum_{i=1}^{t} r_i$. Ergo, $\Sigma$ is
generated by elements whose first coordinate is less than
$\sum_{i = 1}^{t} r_i$.
\end{proof}

@@

Let $M$ be the set of monomials of $\beta_j$.  Then for $\mu\in M$, the corresponding lattice point is of the form
\[
	E_\mu + \sum_{k\in \Te} r_k e_k + \sum_{\substack{
	h \in \Tp \\ k \in \Tm}} r_{h,k} e_{h,k}
\]
where $E_\mu$ is a $\mathbb{R}_{\ge 0}$-combination of elements in
\begin{align*}
	\{e_j : j \in \{1, 2, 3, 4\} \cap \Te\}
	&\cup \; \{e_{j, k}: k \in \Tm, \; j \in \{1, 2, 3, 4\} \cap \Te\} \\
	&\cup \; \{e_{k, j}: k \in \Tp, \; j \in \{1, 2, 3, 4\} \cap \Tm\}.
\end{align*}

Let $F \in R_D$ correspond to the lattice point
\[
	E + \sum_{k \in \Te} \fr(r_k) e_k + \sum_{\substack{h \in \Tp \\
	k \in \Tm}} \fr(r_{h,k}) e_{h,k}
\]

\noindent
where $\fr(r)$ signifies the fractional part of $r$.  Similarly, let $F_\mu
\in R_D$ correspond to the lattice point
\[
	E_\mu + \sum_{\substack{h \in \Tp \\ k \in \Tm}} \fr(r_{h,k}) e_{h,k}.
\]

\noindent
Then $F - \sum_{\mu \in M} F_\mu = 0$ gives us a
relation in degree at most
\[
	\sigma + \deg(F)
\]

\noindent
which divides $(z_i - \beta_i) \prod_{i = 1}^n y_i^{c_i}$.
We can bound $\deg(F)$ by
\[
	\deg(F) \le \begin{cases}
	\ell_i \gcd(\bida_i, \bidb_i)	&\mbox{ if } i \in S_= \\
	\max_{j \in S_-} \bigl(\ell_{i, j} (\bida_i \bidb_j - \bida_j \bidb_i)\bigr)
	&\mbox{ if } i \in S_+ \\
	\max_{j \in S_+} \bigl(\ell_{j, i} (\bida_j \bidb_ i - \bida_i \bidb_j) \bigr)
	&\mbox{ if } i \in S_- \end{cases}.
\]

\noindent
Therefore, $\ker(\psi)$ is generated in degree at most
\[
	\tau = \sigma
	+ \max \left(\max_{i\in S_=} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\; \max_{\substack{i \in S_+ \\ j \in S_-}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j
	\bidb_i) \bigr) \right).
\]
\end{proof}

@@

\noindent
where
\[
	\tau = \sigma
	+ \max \left( \max_{i\in S_=} \bigl(\ell_i \gcd(a_i, b_i) \bigr),
	\max_{\substack{i \in S_+ \\ j \in S_-}} \bigl(\ell_{i, j}
	(\bida_i \bidb_j - \bida_j \bidb_i) \bigr) \right).
\]
\end{thm}
